|

### 2022-05-17 article/article.Rmd
copyediting files

Legislative institutions are crucial for citizen welfare and public service
provision in democratic countries. Historically, the separation of powers and
the birth of representative governments placed legislatures at the center of
the political stage. In modern democracies, most countries have assemblies at
all levels of government. These institutions regulate the quantity and the
quality of legislation, check and veto other political powers, and influence
policymaking. Moreover, legislatures also determine taxation, government
spending and oversee public programs \citep{auriol2012optimal,
north1989constitutions, poulsen2018political, stasavage2011states,
Weingastetal1981}.

Regardless of eventual differences in electoral rules, social composition, and
internal procedures, all legislatures share one common feature: They are
collective bodies. As such, the number of elected representatives has important
consequences for the whole polity. Legislature size is known to have a
significant effect on collective action strategies, the bargaining power
between state branches, government spending, and representation
\citep{crain1979cost, freire2021legislature, gerring2020population,
PrimoSnyder2008, rogers2002free, Weingastetal1981}. However, how legislature
size impacts social welfare remains less understood. On the one hand, enlarged
legislatures may reduce political particularism and increase overall government
spending, which can lead to more public goods \citep{gerring2020population,
GilliganMatsusaka2001, Weingastetal1981}. On the other hand, adding more
representatives to a legislature may exacerbate collective action problems,
thus hindering service provision \citep{gottlieb2019countervailing,
tsebelis2002veto}. Therefore, what is the effect of larger legislatures on
public service delivery?

@@

This paper incorporates partisanship and mayoral first-move advantage to a
\citet{baron1989bargaining} bargaining model. Mayors make the initial policy
proposal, which consists of investments in public goods provision, rents for
personal gains, and a vector of side-transfers for councilors. If mayors fail
to secure majoritarian support from councilors, a reversal stage occurs, in
which the decision leaves the mayor's office and councilors propose the public
investments. In this mayor-council bargaining process, costs come from the
side-transfers required to secure a majority among councilors. We show that in
a model without partisan concerns, bargaining costs always increase with larger
city councils, which in turn lower public goods provision and rents. However,
partisanship decreases bargaining costs when the chances of electing
government-aligned councilors are sufficiently high. Lower bargaining costs
release funds for mayors to invest in public goods. In this case, a larger city
council may add more members to the government coalition, thus reducing the
mayor's bargaining costs and increasing the equilibrium public goods provision.

We test these theoretical predictions by exploiting exogenous variation in city
council size in Brazil. In 2004, the judiciary reinterpreted the Brazilian
Constitution and unexpectedly changed the number of councilors in several
municipalities. Until then, cities could freely determine their legislature
size, but in March 2004 the Supreme Electoral Court created population
thresholds and added one councilor to each set of 47,619 inhabitants. Around
these cutoffs, the decision represented an exogenous change in city council
size, allowing us to investigate the effects of increasing the number of local
legislators in Brazil.

We find that larger city councils changed the composition of mayoral coalitions
and the number of appointed bureaucrats in the municipality. The extra city
councilor had a 91 percent chance of belonging to the mayor's pre-electoral
coalition. With a larger coalition, mayors face lower bargaining costs and are
able to boost public investments with fewer constraints. Each additional
legislator also leads to 104 new appointed bureaucrats, and research shows that
political appointees improve public service provision and enhance
accountability in Brazil \citep{toral2019benefits}.

@@

This paper contributes to four strands of the literature. First, this paper
dialogues with a large literature on multilateral bargaining. Repeated models
of multilateral bargaining started with \citet{baron1989bargaining}, who
extended the bilateral bargaining game proposed by
\citet{rubinstein1982perfect} to more than two players. Several papers followed
\citet{baron1989bargaining} and included endogenous recognition power, the size
of the required majority, and partisanship in the model
\citep{agranov2020persistence, baranski2020determinants,
calvert2005legislative, choate2019partisan}.\footnote{See
\citet{eraslan2019legislative} for a comprehensive review of multilateral
bargaining and \citet{ray2015coalition} for a review on coalition formation.}
In this article, we extend the \citet{baron1989bargaining} model under closed
rules and add the mayor's first move and political partisanship. We consider
partisanship as a zero-sum effect that is positive for politicians aligned with
the mayor and negative otherwise. We investigate the often overlooked effect of
legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature on the effects of larger
legislatures in service provision. Following the seminal contribution by
\citet{Weingastetal1981}, the "law of $1/n$" posits that larger legislatures
increase the legislative pressure for providing targeted public goods to local
supporters at the expense of the entire polity \citep{freire2021legislature}.
This question has become more relevant since legislature size has increased
consistently over time \citep{gerring2020population}. The corollary of the "law
of $1/n$" is that increasing legislature size promotes excessive public
spending, regardless of the partisan makeup of the legislature
\citep{GilliganMatsusaka2001}.\footnote{\citet{PrimoSnyder2008} show that
legislature size may result in positive or negative expenditures, depending on
factors such as the degree that local public goods can exclude non-locals or
deadweight costs associated with taxation.} In the US metaphor, this would
explain significant infrastructure spending in smaller districts. However, the
literature on the "law of $1/n$" rarely mentions the relationship between
public goods provision and welfare. This paper is one of the first to study how
the number of local legislators affects citizen well-being. Our model provides
a theoretical expectation based on the chances of electing a politician aligned
with the mayor. The empirical estimates confirm our theoretical predictions,
with the additional fact that legislature size has a null effect on
expenditures in Brazil.

Third, our article also contributes to the growing literature on municipal
partisan politics. Most papers in this field estimate the effects of the
mayor's party on political outcomes. For instance, \citet{de2016mayoral} and
\citet{gerber2011mayors} find differences in governance when comparing
Democratic and Republican mayors. Specifically about Brazil,
\citet{gouvea2021partisanship} show that social policy expenditure increases
with left-wing mayors. \citet{frey2021legislature}, using discontinuities in
city council size after the 2009 constitutional amendment that reverted the
thresholds we study here, shows that changes in council size lower the support
for candidates backed by the municipal mayors. Our paper reinforces the view
that partisanship matters locally, even in countries with weak ideological
linkage.

Finally, this paper contributes to the literature on the causes and effects of
legislature sizes. While most papers on legislature size analyse the trade-off
between representation and efficiency \citep{frederick2008people,
jacobs2015explaining, muzzio1989size, Stigler1976}, recent articles have
explored the effects of legislature size on lawmaking and partisanship in
further detail. \citet{kirkland2014chamber} shows that larger legislatures make
it harder for politicians to know and collaborate with each other, so they rely
more heavily on party cues and committee structures. In considering the causes
of changes in legislature sizes, \citet{jacobs2015explaining} show that
population shifts increase and economic crises decrease the number of
legislators. \citet{marland2019fewer} questions the economic crises mechanism,
showing that political elites shrink legislatures mostly to signal to the
public the priority of their austerity agenda. In this article, we show that
council size affects welfare by changing the partisan composition of
the legislatures.

@@

Institutional design has a major impact on long-term welfare
\citep{acemoglu2012nations, buenodemesquita2005logic, north1989constitutions}
and democratic consolidation \citep{gandhi2019institutional,
lijphart2018institutional}. One particular problem that all polities face is
how to determine the optimal size of their legislative bodies. The literature
on this topic is extensive, and modern attempts to provide a solution to the
legislature size problem started in the early eighteenth century with
the Federalists versus Anti-Federalists debates \citep{muzzio1989size,
Stigler1976}.

The literature on legislature size highlights that the number of legislators
causes tension between representation and efficiency
\citep{frederick2008people, jacobs2015explaining, muzzio1989size,
rush2013conclusion, Stigler1976, taagepera1972size}. Larger legislatures
improve representation as both the election thresholds and the ratio of
population-per-politician decrease, which facilitate the electoral success of
ethnic and political minorities \citep{allenrepresentation,
gerring2020population}. However, better representation is often countered by
the decline in efficiency in larger legislatures \citep{Weingastetal1981}. The
enlarged size makes it harder for members to coordinate and collaborate without
partisanship and legislative structures \citep{kirkland2014chamber,
rush2013conclusion}. It also increases the potential number of veto players,
decreasing policymaking efficiency \citep{tsebelis2002veto}. Balancing these
two incentives is crucial for the optimal legislative composition
\citep{auriol2012optimal, taagepera1972size}.

This literature has often overlooked two important questions. First, does
legislature size matter in contexts where legislative bodies are not the
pivotal decision-maker? And second, what is the ultimate effect of legislature
size on welfare? These gaps are likely generated by the predominance of the
"law of 1/n" scholarship in political economy debates, which focus mainly on
the relationship between legislature size and overspending. Since
\citet{Weingastetal1981}, 30 papers have systematically tested the effects of
legislature size on government expenditures \citep{freire2021legislature}. Of
these papers, only \citet{lewis2019legislature} investigates its effects on
welfare, showing that larger legislatures have a negative welfare
effect.\footnote{\citet{koppel2004public} also shows, using a repeated game
where the proposal prerogative rotates in a legislature, that larger
legislatures increase the chances of non-cooperative outcomes, lowering public
goods provision.} \citet{frey2021legislature} analyzes the Brazilian case after
the 2009 legislature size change and finds a null effect in service provision
from 2012 to 2016. Yet, \citet{GilliganMatsusaka2001} find that in the US
legislature size is associated primarily with higher spending on geographically
concentrated benefits, such as education and highways. These results suggest
that the legislature size and welfare nexus needs a dedicated theoretical and
empirical assessment.

@@

In a mayor-council setting, consider a strategic interaction between a mayor
($M$) and $N \geq 3$ city councilors indexed in $i \in \{1, 2, \cdots, N \}$.
The mayor has the prerogative of proposing a vector of policies on which city
councilors vote. The provision of public goods is then a bargain between the
mayor and the city council. Policy proposals are a combination of public goods
provision $g$, rents for the mayor $r$, and a vector of transfer for city
councilors \textbf{x}.

The city council votes the mayor's proposal, and if councilors accept it, the
proposal is implemented. Otherwise, a reversal policy takes place. Reversal
policy in this context means that the mayor's proposal failed to secure the
majority of council votes. The decision process then leaves the mayor's office
and is transferred to the council. The types of reversal policies are crucial
for our argument, as they change the relative strength of the council when
legislators bargain with the mayor. We investigate two types of reversal
policies: A baseline reversal with no parties and a hybrid mechanism with
partisan and non-partisan concerns. The differences between these two
mechanisms are that the non-partisan councilors only care about rents, while
partisan councilors care about both rents and policy.

The mayor must convince at least half of the city councilors to support her
policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that
the selectorate, that is, the individuals who have a say in the mayor's policy
proposals, are the $N$ city councilors. The winning coalition, the minimum
number of members that must be convinced to implement the policy, is equal to
half of the selectorate ($N/2$). City councilors are motivated by policy
concerns $p$ and transfers received from the mayor or the councilor nominated
as the proposer \textbf{x}. We leave the definition of these transfers open, as
this allows us to analyze how mayors combine transfers with other incentives to
pressure city councilors. Exemples of transfers are granting electoral favors,
funding clientelistic brokers, spending in areas that the councilor has
political or personal interests, providing portfolio (municipal offices) and
public jobs within the municipal bureaucracy, or even giving campaign
contributions and bribes. The use of transfers to influence the councilors'
choices generate a bargaining cost of $C_G$. Bargaining costs vary according to
the reversal mechanism and the size of the legislature.

Finally, after learning the bargaining costs, the mayor provides a level of
public goods provision $g$ and pockets rents $r$. Public goods provision helps
the mayor to get reelected. Rents are for the mayor's direct consumption and do
not contribute to the mayor's electoral success.\footnote{This model, similar
to \citet{cowen1994rent} and \citet{buenodemesquita2005logic}, combines public
goods provision and rent extraction. For a dynamic bargaining model that
incorporates transfers to districts and public goods provision, see
\citet{battaglini2012legislative}.} This makes the mayor's expected utility a
sum of the gains from rents and the benefits from reelection. Both the
utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are
concave functions, meaning that more rents or public goods increase the utility
at a decreasing rate. The probability of reelection is multiplied by the
benefit from holding office $B_M > 0$. This benefit captures the tangible and
intangible gains that the mayor receives from holding the public office. We
assume that the benefits are high enough to rule out an equilibrium with only
rents and transfers, which is equivalent to saying that mayors strongly prefer
to be reelected. If the mayor's policy proposal is approved, then the expected
utility for the mayor is:

@@

We now consider a situation that combines both partisan and non-partisan
legislature concerns. Each councilor has a party affiliation. Party
affiliations are mutually exclusive, and if a city councilor is aligned with
the mayor's party, then she belongs to the government coalition $G \subset \{1,
2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O =
\{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political
value of $p > 0$ for the councilors, and whether it adds or subtracts from the
councilor's utility depends on the party affiliation of the
councilor.\footnote{One example of partisan benefits is that, as the mayor
shares partisan ties with members of the government coalition, a fraction of
the public goods provision can be claimed by councilors as their contribution.
We propose an extension in the Online Supplementary Materials, which assumes $p
= \phi g$, where $\phi$ is the fraction of the public goods provision that can
be credit claimed by the city councilor aligned with the mayor.} If councilors
reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in
equilibrium, we assume that city councilors would prefer to receive transfers
instead of providing public goods.\footnote{In the Online Supplementary
Materials, we consider a version of the model in which public goods provision
depends on the party affiliation of the proposer so that it affects choices in
any potential history of the game. In equilibrium, the level of public goods
provision is even higher than in the model presented here, as city councilors
prefer to approve the mayor's proposal than run the risk of an opposition
member becoming the proposer.}

@@

\begin{enumerate}
  \item The mayor learns how many government $|G|$ and opposition $|O|$ legislators were elected.
  \item The mayor proposes a policy vector $(r, g, x)_M$.
  \item The city council votes the proposal.
  \begin{itemize}
    \item If the council accepts the proposal, the policy is implemented and the game ends. 
    \item Otherwise, the reversal policy is implemented.
  \end{itemize}
  \item (Reversal Policy) The budget is discounted by a factor $\delta \in (0,1)$, and one councilor is randomly selected to make an offer. If the offer is accepted, it is implemented. If rejected, the reversal policy stage restarts.\footnote{In section A.1 of the Supplementary Materials, we solve three reversal mechanisms. We include the two mechanisms we present here, plus a third mechanism with only political concerns.}
\end{enumerate}

@@

All proofs are in section A.1 of the Supplementary Materials. The proof's
intuition involves finding what happens to the mayor's marginal utility when we
increase both $g$ and $N$. As $N$ increases discretely on a complete lattice,
we use monotone comparative statics to derive the conditions for increasing
differences \citep{milgrom1994monotone, ashworth2006monotone}. Increasing
differences mean that higher $N$ leads to higher equilibrium $g$.

@@

We place the offer on the left-hand side. There are two components on the
right-hand side. The first is the amount that the councilor $i$ gets when she
is the proposer. It is equal to the chance that she is recognized as the
proposer times the budget in the next round minus the offers she makes to
convince half of the councilors. The second part represents the gains if the
councilor rejects the current received offer but still gets an offer in the
following round. It is equal to the chance that she is not recognized as the
proposer times the chance that she receives a transfer times the transfer
amount. Note that the offer the councilor makes as the proposer is the same as
the offer she wants to receive. This is because the city councilors are
exchangeable. The solution is symmetric for all councilors receiving an offer
(this means that, without loss of generality, we could have dropped the $i$ in
the solution). After some algebra, the offer $x_i$ must be greater than or
equal to:

@@

The intuition for the results is that every additional councilor demands more
resources from the mayor. Thus, more councilors represent higher bargaining
costs. And if the bargaining costs increase with legislature size, then any
increase in public goods provision associated with legislature size has come
from a different mechanism. Below we demonstrate that partisanship is the key
to understand how this relationship works.

@@

We decompose these costs into two components. First, the costs in terms of
rents. Second, the costs (or benefits) from political alignment. In this
context, the bargaining costs depend on the mayor's offers for councilors in
her coalition versus the mayor's proposals for councilors in the opposition. As
the ex-ante chance of an opposition member being elected is equal to
$1-\gamma$, after taking the weighted averages these costs become:

@@

The intuition for the proof is that as partisanship influences the costs and
benefits of legislators, more government-aligned councilors make it easier for
the mayor to govern. When the partisan benefit is higher than the additional
cost that a legislator generates, adding a councilor will benefit mayors if the
councilor has a higher chance of sharing the same party as the mayor.
Therefore, a higher probability of electing a legislator aligned with the mayor
reverts into lower bargaining costs.

@@

In 2004, Brazil comprised 5,560 municipalities. According to the Brazilian
Constitution, each city must provide health care, primary education,
transportation, and infrastructure to its citizens. Municipalities have the
authority to enact local laws and collect taxes on housing and services.
However, their room for fiscal manipulation is limited. On average, municipal
taxes account for only 5\% of the local budgets, while mandatory transfers from
state and federal governments correspond to the remaining 95\%. This reduces
the ability of municipal legislators to effectively increase local spending,
which makes improvements in local services dependent only on the composition of
their local investments.

@@

Until 2004, there was no specific regulation on the size of city councils. The
1988 Brazilian Constitution set only broad guidelines about the number of
council seats in each municipality, allowing cities to determine how many
legislators their local chambers should have. This led to severe imbalances in
the representation ratio across the country. For instance, the Nova Russas
municipality had 30,009 inhabitants and 21 councilors, or one legislator per
1,429 citizens. In contrast, the city of Sorocaba had only 15 councilors for
528,735 inhabitants, that is, one legislator per 35,249 citizens. In 2003, this
malapportionment problem gained national visibility when Mira Estrela, a small
municipality with only 2,651 residents, reduced its city council from 11 to 9
seats because of fiscal considerations. The change motivated a legal dispute
that reached the Brazilian Supreme Court. The Court favored the city's decision
and ruled that the size of every local council should correspond to precise
population thresholds.

>>>>

Legislative institutions are crucial for citizen welfare and public service
provision in democratic countries. Historically, the separation of powers and
the birth of representative governments placed legislatures at the center of
the political stage. In modern democracies, most countries have assemblies at
all levels of government. These institutions regulate the quantity and quality
of legislation, check and veto other political powers, and influence
policymaking. Moreover, legislatures also determine taxation, regulate
government spending, and oversee public programs \citep{auriol2012optimal,
north1989constitutions, poulsen2018political, stasavage2011states,
Weingastetal1981}.

Regardless of eventual differences in electoral rules, social composition, or
internal procedures, all legislatures share one common feature: They are
collective bodies. As such, the number of elected representatives has important
consequences for the whole polity. Legislature size is known to have a
significant effect on the bargaining power between state branches, public
expenditures, and representation \citep{crain1979cost, freire2021legislature,
gerring2020population, PrimoSnyder2008, rogers2002free, Weingastetal1981}.
However, how legislature size impacts social welfare remains less understood.
On the one hand, larger legislatures may reduce political particularism and
increase overall government spending, which can lead to more public goods
\citep{gerring2020population, GilliganMatsusaka2001, Weingastetal1981}. On the
other hand, adding representatives to a legislature may exacerbate collective
action problems, thus hindering service provision
\citep{gottlieb2019countervailing, tsebelis2002veto}. Therefore, what is the
effect of larger legislatures on public service delivery?

@@

This paper incorporates partisanship and mayoral first-move advantage to a
\citet{baron1989bargaining} bargaining model. Mayors make the initial policy
proposal, which consists of investments in public goods provision, rents for
personal gains, and a vector of side-transfers for councilors. If mayors fail
to secure majoritarian support from councilors, a reversal stage occurs, in
which the decision leaves the mayor's office and councilors propose the public
investments. In this bargaining process, costs come from the side-transfers
required to secure a majority among councilors. We show that in a model without
partisan concerns, bargaining costs always increase with larger city councils,
which in turn lower public goods provision and rents. However, partisanship
decreases bargaining costs when the chances of electing government-aligned
councilors are sufficiently high. Lower bargaining costs release funds for
mayors to invest in public goods. In this case, a larger city council may add
more members to the government coalition, thus reducing the mayor's bargaining
costs and increasing the equilibrium public goods provision.

We test these theoretical predictions by exploiting exogenous variation in city
council size in Brazil. In 2004, the judiciary reinterpreted the Brazilian
Constitution and unexpectedly changed the number of councilors in several
municipalities. Until then, cities could freely determine their legislature
size. But in March 2004 the Supreme Electoral Court created population
thresholds and added one councilor to each set of 47,619 inhabitants. Around
these cutoffs, the decision represented an exogenous change in city council
size, allowing us to investigate the effects of increasing the number of local
legislators in Brazil.

We find that larger city councils changed the composition of mayoral coalitions
and the number of appointed bureaucrats in the municipality. The extra city
councilor had a 91 percent chance of belonging to the mayor's pre-electoral
coalition. With a larger coalition, mayors face lower bargaining costs and are
able to boost public investments with fewer constraints. Each additional
legislator also leads to 104 politically appointed bureaucrats, and research
shows that political appointees improve public service provision and enhance
accountability in Brazil \citep{toral2019benefits}.

@@

This paper contributes to four strands of the literature. First, this paper
dialogues with a large literature on multilateral bargaining. Repeated models
of multilateral bargaining started with \citet{baron1989bargaining}, who
expanded the bilateral bargaining game proposed by
\citet{rubinstein1982perfect} to more than two players. Several authors then
analysed the effect of endogenous recognition power, the size of the required
majority, and partisanship on these models \citep{agranov2020persistence,
baranski2020determinants, calvert2005legislative,
choate2019partisan}.\footnote{See \citet{eraslan2019legislative} for a
comprehensive review of multilateral bargaining and \citet{ray2015coalition}
for a review on coalition formation.} In this article, we extend
\citeauthor{baron1989bargaining}'s (\citeyear{baron1989bargaining}) model under
closed rules and add the mayor's first move and political partisanship. We
consider partisanship as a zero-sum effect that is positive for politicians
aligned with the mayor and negative otherwise. We investigate the often
overlooked effect of legislature size on the bargaining equilibrium strategies.

Second, this paper contributes to the literature on the effects of larger
legislatures on service provision. Following the seminal contribution by
\citet{Weingastetal1981}, the "law of $1/n$" posits that larger chambers
provide incentives for legislators to provide targeted public goods to local
supporters at the expense of the entire polity \citep{freire2021legislature}.
This question has become more relevant since legislature size has increased
consistently over time \citep{gerring2020population}. The corollary of the "law
of $1/n$" is that increasing legislature size promotes excessive public
spending, regardless of the partisan makeup of the legislature
\citep{GilliganMatsusaka2001}.\footnote{\citet{PrimoSnyder2008} show that
legislature size may result in positive or negative expenditures, depending on
factors such as the degree that local public goods can exclude non-locals or
deadweight costs associated with taxation.} In the US case, this would
explain significant infrastructure spending in smaller districts. However, the
literature on the "law of $1/n$" rarely mentions the relationship between
public goods provision and welfare. This paper is one of the first to study how
the number of local legislators affects citizen well-being. Our model provides
a theoretical expectation based on the chances of electing a politician aligned
with the mayor. The empirical estimates confirm our theoretical predictions,
with the additional fact that legislature size has a null effect on
expenditures in Brazil.

Third, our article also dialogues with the growing literature on municipal
partisan politics. Most papers in this field estimate the effects of the
mayor's party on political outcomes. For instance, \citet{de2016mayoral} and
\citet{gerber2011mayors} find differences in governance when comparing
Democratic and Republican mayors. Specifically about Brazil,
\citet{gouvea2021partisanship} show that social spending increases with
left-wing mayors. \citet{frey2021legislature}, using discontinuities in city
council size after the 2009 constitutional amendment that reverted the
thresholds we study here, shows that changes in council size lower the support
for candidates backed by the municipal mayors. Our paper reinforces the view
that partisanship matters locally, even in countries with weak ideological
linkage.

Finally, this paper contributes to the literature on the causes and effects of
legislature sizes. While most papers on legislature size analyse the trade-off
between representation and efficiency \citep{frederick2008people,
jacobs2015explaining, muzzio1989size, Stigler1976}, recent articles have
explored the effects of legislature size on lawmaking and partisanship in
further detail. \citet{kirkland2014chamber} shows that larger legislatures make
it harder for politicians to know and collaborate with each other, so they rely
more heavily on party cues and committee structures. In considering the causes
of changes in legislature sizes, \citet{jacobs2015explaining} show that while
population shifts tend to increase the number of local legislators, economic
crises decrease them. \citet{marland2019fewer} questions the latter mechanism,
showing that political elites shrink legislatures mostly to signal to the
public the priority of their austerity agenda. In this article, we show that
council size affects welfare by changing the partisan composition of the
legislatures.

@@

Institutional design has a lasting impact on social welfare
\citep{acemoglu2012nations, buenodemesquita2005logic, north1989constitutions}
and democratic consolidation \citep{gandhi2019institutional,
lijphart2018institutional}. One particular problem that all polities face is
how to determine the optimal size of their legislative bodies. The literature
on this topic is extensive, and modern attempts to address the legislature size
problem started in the early eighteenth century with the Federalists versus
Anti-Federalists debates \citep{muzzio1989size, Stigler1976}.

The literature on legislature size highlights that the number of legislators
causes tension between representation and efficiency
\citep{frederick2008people, jacobs2015explaining, muzzio1989size,
rush2013conclusion, Stigler1976, taagepera1972size}. Larger legislatures
improve representation as both the election thresholds and the ratio of
population-per-politician decrease, which facilitate the electoral success of
ethnic and political minorities \citep{allenrepresentation,
gerring2020population}. However, better representation is often countered by
the decline in efficiency in larger legislatures \citep{Weingastetal1981}. In
large council, it is harder for members to collaborate with each other without
some type of partisanship or formal legislative structure
\citep{kirkland2014chamber, rush2013conclusion}. Larger chambers also have more
potential veto players, who may decrease policymaking efficiency
\citep{tsebelis2002veto}. In this sense, balancing representation and
efficiency is key to find the optimal legislative composition
\citep{auriol2012optimal, taagepera1972size}.

However, two questions have often been overlooked in these debates. First, does
legislature size matter in contexts where legislative bodies are not pivotal
decision-makers? Second, what is the ultimate effect of legislature size on
welfare? These gaps are likely generated by the predominance of the "law of
1/n" scholarship in this literature, which focuses mainly on the relationship
between legislature size and overspending. Since \citet{Weingastetal1981}, 30
papers have tested the effects of legislature size on government expenditures
\citep{freire2021legislature}, and the only article which examined its effects
on welfare found negative results
\citep{lewis2019legislature}.\footnote{\citet{koppel2004public} uses a repeated
game where the proposal prerogative rotates in a legislature and finds that
larger legislatures increase the chances of non-cooperative outcomes, lowering
public goods provision.} \citet{frey2021legislature} analyzes the Brazilian
case after the 2009 legislature size change and finds a null effect in service
provision from 2012 to 2016. Yet, \citet{GilliganMatsusaka2001} find that in
the US legislature size is associated primarily with higher spending on
geographically concentrated benefits, such as education and highways. These
results suggest that the legislature size and welfare nexus needs a dedicated
theoretical and empirical assessment.

@@

In a mayor-council setting, consider a strategic interaction between a mayor
($M$) and $N \geq 3$ city councilors indexed in $i \in \{1, 2, \cdots, N \}$.
The mayor has the prerogative of proposing a vector of policies to be voted by
the city councilors. The provision of public goods is then a bargain between
the mayor and the city council. Policy proposals are a combination of public
goods provision $g$, rents for the mayor $r$, and a vector of transfer for city
councilors \textbf{x}.

The city council votes the mayor's proposal and the proposal is implemented if
most councilors accept it. Otherwise, a reversal policy takes place. Reversal
policy in this context means that the mayor's proposal failed to secure the
majority of council votes. The decision process then leaves the mayor's office
and is transferred to the council. The types of reversal policies are crucial
for our argument, as they change the relative strength of the council when
legislators bargain with the mayor. We investigate two types of reversal
policies: A baseline reversal with no parties and a hybrid mechanism with
partisan and non-partisan concerns. The differences between these two
mechanisms are that the non-partisan councilors only care about rents, while
partisan councilors care about both rents and policy.

The mayor must convince at least half of the city councilors to support her
policy proposals. Similar to \citet{buenodemesquita2005logic}, we assume that
the selectorate, that is, the individuals who have a say in the mayor's policy
proposals, are the $N$ city councilors. The winning coalition, which is the
minimum number of members that must be convinced to implement the policy, is
equal to half of the selectorate ($N/2$). City councilors are motivated by
policy concerns $p$ and transfers received from the mayor or the councilor
nominated as the proposer \textbf{x}. We leave the definition of these
transfers open, as this allows us to analyze how mayors combine transfers with
other incentives to pressure city councilors. Exemples of transfers are
granting electoral favors, funding clientelistic brokers, spending in areas in
which the councilor has political or personal interests, or providing portfolio
(municipal offices) and public jobs within the municipal bureaucracy. The use
of transfers to influence the councilors' choices generate a bargaining cost of
$C_G$. Bargaining costs vary according to the reversal mechanism and the size
of the legislature.

Finally, after learning the bargaining costs, the mayor provides a level of
public goods provision $g$ and pockets rents $r$. Public goods provision helps
the mayor to get reelected. Rents are for the mayor's direct consumption and do
not contribute to her electoral success.\footnote{This model, similar to
\citet{cowen1994rent} and \citet{buenodemesquita2005logic}, combines public
goods provision and rent extraction. For a dynamic bargaining model that
incorporates transfers to districts and public goods provision, see
\citet{battaglini2012legislative}.} This makes the mayor's expected utility a
sum of the gains from rents and the benefits from reelection. Both the
utilities from rents $u(.)$ and the probability of reelection $\pi(.)$ are
concave functions, meaning that more rents or public goods increase the utility
at a decreasing rate. The probability of reelection is multiplied by the
benefit from holding office $B_M > 0$. This benefit captures the tangible and
intangible gains that the mayor receives from holding the public office. We
assume that the benefits are high enough to rule out an equilibrium with only
rents and transfers, which is equivalent to saying that mayors strongly prefer
to be reelected. If the mayor's policy proposal is approved, then the expected
utility for the mayor is:

@@

We now consider a situation that combines both partisan and non-partisan
legislature concerns. Each councilor has a party affiliation. Party
affiliations are mutually exclusive, and if a city councilor is aligned with
the mayor's party, then she belongs to the government coalition $G \subset \{1,
2, \cdots, N \}$. Otherwise, the city councilor belongs to the opposition $O =
\{1, 2, \cdots, N \} \setminus G$. The mayor's policy generates a political
value of $p > 0$ for the councilors, and whether it adds or subtracts from the
councilor's utility depends on the party affiliation of the
councilor.\footnote{One example of partisan benefits is that, as the mayor
shares partisan ties with members of the government coalition, a fraction of
the public goods provision can be claimed by councilors as their contribution.
We propose an extension in the Supplemental Materials, which assumes $p = \phi
g$, where $\phi$ is the fraction of the public goods provision that can be
credit claimed by the city councilor aligned with the mayor.} If councilors
reject the mayor's proposal, no public goods are provided ($p=0$). Moreover, in
equilibrium, we assume that city councilors would prefer to receive transfers
instead of providing public goods.\footnote{In the Supplemental Materials, we
consider a version of the model in which public goods provision depends on the
party affiliation of the proposer so that it affects choices in any potential
history of the game. In equilibrium, the level of public goods provision is
even higher than in the model presented here, as city councilors prefer to
approve the mayor's proposal than run the risk of an opposition member becoming
the proposer.}

@@

\begin{enumerate}
  \item The mayor learns how many government $|G|$ and opposition $|O|$ legislators were elected.
  \item The mayor proposes a policy vector $(r, g, x)_M$.
  \item The city council votes the proposal.
  \begin{itemize}
    \item If the council accepts the proposal, the policy is implemented and the game ends. 
    \item Otherwise, the reversal policy is implemented.
  \end{itemize}
  \item (Reversal Policy) The budget is discounted by a factor $\delta \in (0,1)$, and one councilor is randomly selected to make an offer. If the offer is accepted, it is implemented. If rejected, the reversal policy stage restarts.\footnote{In section A.1 of the Supplemental Materials, we solve three reversal mechanisms. We include the two mechanisms we present here, plus a third mechanism with only political concerns.}
\end{enumerate}

@@

All proofs are in section A.1 of the Supplemental Materials. The proof's
intuition involves finding what happens to the mayor's marginal utility when we
increase both $g$ and $N$. As $N$ increases discretely on a complete lattice,
we use monotone comparative statics to derive the conditions for increasing
differences \citep{milgrom1994monotone, ashworth2006monotone}. Increasing
differences mean that higher $N$ leads to higher equilibrium $g$.

@@

We place the offer on the left-hand side. There are two components on the
right-hand side. The first is the amount that the councilor $i$ gets when she
is the proposer. It is equal to the chance that she is recognized as the
proposer times the budget in the next round minus the offers she makes to
convince half of the councilors. The second part represents the gains if the
councilor rejects the current offer but still gets an offer in the following
round. It is equal to the chance that she is not recognized as the proposer
times the chance that she receives a transfer times the transfer amount. Note
that the offer the councilor makes as the proposer is the same as the offer she
wants to receive. This is because city councilors are exchangeable. The
solution is symmetric for all councilors receiving an offer (this means that,
without loss of generality, we could have dropped the $i$ in the solution).
After some algebra, the offer $x_i$ must be greater than or equal to:

@@

The intuition for the results is that every additional councilor demands more
resources from the mayor. Thus, more councilors represent higher bargaining
costs. And if the bargaining costs increase with legislature size, then any
increase in public goods provision associated with legislature size has to come
from a different mechanism. Below we demonstrate that partisanship is the key
to understand how this relationship works.

@@

We decompose these costs into two components, which are the costs in terms of
rents and the costs (or benefits) from political alignment. In this context,
the bargaining costs depend on what the mayor offers for councilors in her
coalition versus what she provides to members of the opposition. As the ex-ante
chance of an opposition member being elected is equal to $1-\gamma$, after
taking the weighted averages these costs become:

@@

The intuition for the proof is the following. As partisanship influences the
costs and benefits of legislators, more government-aligned councilors make it
easier for the mayor to govern. When the partisan benefit is higher than the
additional cost that a legislator generates, adding a councilor will benefit
mayors if the councilor has a higher chance of sharing the same party as the
mayor. Therefore, a higher probability of electing a legislator aligned with
the mayor reverts into lower bargaining costs.

@@

In 2004, Brazil comprised 5,560 municipalities. According to the 1988 Brazilian
Constitution, each city must provide health care, primary education,
transportation, and infrastructure to its citizens. Municipalities have the
authority to enact local laws and collect taxes on housing and services.
However, their room for fiscal manipulation is limited. On average, municipal
taxes account for only 5\% of the local budgets, while mandatory transfers from
state and federal governments correspond to the remaining 95\%. This reduces
the ability of municipal legislators to effectively increase local spending,
which makes improvements in local services dependent only on the composition of
their local investments.

@@

Until 2004, there was no specific regulation on the size of city councils. The
Brazilian Constitution set only broad guidelines about the number of council
seats in each municipality, allowing cities to determine how many legislators
their local chambers should have. This led to severe imbalances in the
representation ratio across the country. For instance, the Nova Russas
municipality had 30,009 inhabitants and 21 councilors, or one legislator per
1,429 citizens. In contrast, the city of Sorocaba had only 15 councilors for
528,735 inhabitants, that is, one legislator per 35,249 citizens. In 2003, this
malapportionment problem gained national visibility when Mira Estrela, a small
municipality with only 2,651 residents, reduced its city council from 11 to 9
seats because of fiscal considerations. The change motivated a legal dispute
that reached the Brazilian Supreme Court. The Court favored the city's decision
and ruled that the size of every local council should correspond to precise
population thresholds.
